---
title: "Alpha_Diversity_ITS"
author: "Yupawadee Galasong"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
### Load packages
```{r}
pacman::p_load(phyloseq, iNEXT, tidyverse, ggpubr, install = FALSE)
```

### Load color and shape preferences
```{r}
source(file = "code/colors.R")

```

### Biodiversity Analysis: OJ dataset
```{r}
load(file = "Data_ITS/PERFect_physeq.RData")

# Create ASV data frame as an input for iNEXT
iNEXT_input_oj <- PERFect_physeq %>% 
  subset_samples(Beverage == "Orange Juice") %>%
  otu_table() %>%
  as.data.frame()

# For reproducibility
set.seed(777)

# Calculate Hill numbers
#q = 0 means total number of taxa ("Effective number of species")
#q = 1 means Number of common taxa (equivalent to exp(Shanon))
#q = 2 means Number of dominant taxa (inverse Simpson)

iNEXT_output_oj <- iNEXT(iNEXT_input_oj, q = c(0,1,2), datatype = "abundance")
```

#### Export iNEXT output as a dataframe suitable for plotting 
```{r}
# Creating a data frame containing diversity measures from iNEXT output
# plus metadata
metadata <- PERFect_physeq@sam_data %>% data.frame() %>%
  rownames_to_column(var = "Assemblage") %>%
  # Add another grouping variable - Thermal Vs Non-thermal Vs No Treatment
    mutate(Thermal = case_when(
    (Treatment == "Pasteurization")~"Thermal",
    (Treatment == "HPP")~"Non-thermal",
    (Treatment == "HPP + DMDC + NG")~"Non-thermal",
    (Treatment == "Pasteurization + DMDC + NG")~"Thermal",
    (Treatment == "Pasteurization + SB + PS")~"Thermal",
    (Treatment == "UV")~"Non-thermal",
    (Treatment == "No Treatment")~"No Treatment")) %>%
  mutate(Tech = case_when(
  Treatment == "Pasteurization" |
    Treatment == "Pasteurization + DMDC + NG" |
    Treatment == "Pasteurization + SB + PS" ~ "Pasteurization",
  Treatment == "HPP" | 
    Treatment == "HPP + DMDC + NG" ~ "HPP",
  .default = "Untreated & UV")) %>% mutate (Tech = as.factor(Tech))

metadata_oj <- metadata %>% filter(Beverage == "Orange Juice")

div_iNEXT_oj <- iNEXT_output_oj$AsyEst %>%
  left_join(., metadata_oj, by = "Assemblage") %>% 
  # Change diversity measures to factor for plotting
  mutate(Diversity = as.factor(Diversity))
```

#### Species richness over shelf-life
```{r}

# Plot Richness over time, points colored based on 'Treatment'
div_iNEXT_oj %>% filter(Diversity == "Species richness") %>%
  ggplot(aes(x = Day, y = Observed))+
  geom_point(aes(color = Treatment), size = 4, alpha = 0.75)+
  scale_shape_manual(values = Beverage_shape)+
  scale_color_manual(values = Treatment_color)+
  theme_classic() + labs(x = "Time (Days)", y = "Species Richness")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.text = element_text(size = 12))

```

#### Rarefaction curve
```{r}
# 1. Extract the observed richness from iNEXT output
obs_rich_oj <- iNEXT_output_oj$iNextEst$size_based %>%
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Observed") %>%
  # Include sample metadata
  left_join(., metadata_oj, by = "Assemblage") 
  

rarefaction_curve_oj <- 
  # 2. Extract richness calculated from rarefied samples
  iNEXT_output_oj$iNextEst$size_based %>% 
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Rarefaction")%>%
  # Include sample metadata
  left_join(., metadata_oj, by = "Assemblage") %>% 
  # 3. Plot rarefaction curve
  ggplot(aes(x = m/1000, y = qD, color = Treatment, group = Assemblage))+
  scale_color_manual(values = Treatment_color) + 
  geom_line()+ 
  geom_point(data = obs_rich_oj, 
             aes(x = m/1000, y = qD, color = Treatment),
             size = 3)+
  labs(x = "Number of Sequences (x1000)", 
       y = "Species Richness")+
  facet_wrap(.~Trial)+
  theme_pubr()


```

### Diversity boxplot
```{r}
# Diversity in orange juice
obs_rich_oj %>%
  ggplot(aes(x = Shelf_Life_Stage, y = qD, color = Shelf_Life_Stage))+
  geom_boxplot(aes(fill = Shelf_Life_Stage), alpha = 0.5)+
  scale_fill_manual(values = Stage_color)+
  scale_color_manual(values = Stage_color)+
  geom_jitter(size = 2.5, alpha = 0.75)+
  theme_classic()+
  labs(x = "Stage of Shelf Life", 
       y = "Species richness", 
       color = "Stage in Shelf Life",
       fill = "Stage in Shelf Life")+
  facet_grid(.~Tech)+ stat_kruskal_test()+
  #theme(axis.text.x = element_text(angle = 45))
  theme_pubr()

# This plot below shows that in non-thermally processed OJ, the richness in spoiled OJ are significantly lower than samples that are not spoiled.
oj_richness <- obs_rich_oj %>%
  ggplot(aes(x = Spoilage, y = qD, color = Spoilage))+
  geom_boxplot(aes(color = "black",fill = Spoilage), 
               outlier.shape = NA)+ 
  scale_fill_manual(values = Spoiled_color)+
  guides(fill = "none")+
  geom_jitter(size = 2.5)+
  scale_color_manual(values = Treatment_color,
                     name = "Treatment")+
  labs(x = NULL, y = "Species richness")+
  facet_grid(.~Tech)+ 
  stat_kruskal_test(label.y = 50)+
  #theme(axis.text.x = element_text(angle = 45))+
  theme_pubr(legend = "none")
```

### Shannon index for species evenness
```{r}
oj_shannon <-div_iNEXT_oj %>%
  filter(Diversity == "Shannon diversity") %>%
  ggplot(aes(x = Spoilage, y = Observed, fill = Spoilage))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = FALSE)+
  geom_jitter(colour = "black",size = 2.5)+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()+labs(x= NULL, y = "Shannon diversity")+
  facet_grid(.~Tech)+ 
  stat_kruskal_test(label.y = 21)+
  theme_pubr(legend = "right")
```

### Simpson diversity measuring the abundant/dominant taxa
```{r}
 oj_simpson <- div_iNEXT_oj %>%
  filter(Diversity == "Simpson diversity") %>%
  ggplot(aes(x = Tech, y = Observed))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = FALSE)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()+labs(x = NULL, y = "Simpson diversity")+
  facet_grid(.~Shelf_Life_Stage)+
  theme(legend.position = "bottom")+
  stat_kruskal_test(label.y = 15)+
  theme_pubr(legend = "bottom")

```

```{r}
dunn.test.its.oj <- div_iNEXT_oj_its %>% group_by(Diversity) %>%
  dunn_test(Observed ~ Tech)
dunn.test.its.oj$p.format <- p_format(dunn.test.its.oj$p.adj, accuracy = 0.01)


```

```{r}
alpha_OJ_ITS <- div_iNEXT_oj_its %>%
  mutate(Diversity = 
           factor(Diversity, 
                  levels = c("Species richness",
                             "Shannon diversity",
                             "Simpson diversity"))) %>%
  ggplot(aes(x = Tech, y = Observed, color = Treatment, group = Tech))+
  geom_point(alpha = 0.8, size = 3)+
  scale_color_manual(values = Treatment_color)+
  labs(x = NULL, y = "Value")+
  facet_grid(.~Diversity)+
  theme_pubr(legend = "none")+
  theme(axis.text.x = element_text(size = 10, angle = 90))+
  stat_pvalue_manual(dunn.test.its.oj, y.position = c(50,55,60), 
                     label = "p.format")

```



### Combining Shannon and Simpson Diversity Plots
```{r}
ggpubr::ggarrange(oj_richness, oj_shannon, oj_simpson,
                  ncol = 1, nrow = 3, 
                  common.legend = TRUE, 
                  legend = "bottom",
                  labels = c("A", "B", "C"))
ggsave(filename = "Figure/alpha_ITS_oj.png", device = "png", height = 8, width = 8, units = "in")
```

### Biodiversity Analysis: Cider Dataset
```{r}
load(file = "Data_ITS/final_physeq.RData")

# Create ASV data frame as an input for iNEXT
iNEXT_input_cd <- PERFect_physeq %>% 
  subset_samples(Beverage == "Cider") %>%
  otu_table() %>%
  as.data.frame()

# For reproducibility
set.seed(777)

# Calculate Hill numbers
#q = 0 means total number of taxa ("Effective number of species")
#q = 1 means Number of common taxa (equivalent to exp(Shanon))
#q = 2 means Number of dominant taxa (inverse Simpson)

# Remove sample(s) with zero counts across ASVs
iNEXT_input_cd<- iNEXT_input_cd[,-c(which(colSums(iNEXT_input_cd) == 0))]

iNEXT_output_cd <- iNEXT(iNEXT_input_cd, q = c(0,1,2), datatype = "abundance")
```

#### Export iNEXT output as a dataframe suitable for plotting 
```{r}
# Creating a data frame containing diversity measures from iNEXT output
# plus metadata
metadata <- final_physeq@sam_data %>% data.frame() %>%
  rownames_to_column(var = "Assemblage") %>%
  # Add another grouping variable - Thermal Vs Non-thermal Vs No Treatment
    mutate(Thermal = case_when(
    (Treatment == "Pasteurization")~"Thermal",
    (Treatment == "HPP")~"Non-thermal",
    (Treatment == "HPP + DMDC + NG")~"Non-thermal",
    (Treatment == "Pasteurization + DMDC + NG")~"Thermal",
    (Treatment == "Pasteurization + SB + PS")~"Thermal",
    (Treatment == "UV")~"Non-thermal",
    (Treatment == "No Treatment")~"No Treatment"))

metadata_cd <- metadata %>% filter(Beverage == "Cider") %>%
  filter(Assemblage != "ITS066")

div_iNEXT_cd <- iNEXT_output_cd$AsyEst %>%
  left_join(., metadata_cd, by = "Assemblage") %>% 
  # Change diversity measures to factor for plotting
  mutate(Diversity = as.factor(Diversity))
```

#### Species richness over shelf-life
```{r}
# Plot Richness over Stage of Shelf-Life, points colored based on 'Treatment', shaped by 'Thermal'
div_iNEXT_cd %>% filter(Diversity == "Species richness") %>%
  ggplot(aes(x = Shelf_Life_Stage, y = Observed))+
  geom_point(aes(color = Treatment, shape = Thermal), size = 4, alpha = 0.75)+
  scale_color_manual(values = Treatment_color)+
  theme_classic() + labs(x = "Stage in Shelf-life", y = "Species Richness")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.text = element_text(size = 12))

# Plot Richness over time, points colored based on 'Treatment'
div_iNEXT_cd %>% filter(Diversity == "Species richness") %>%
  ggplot(aes(x = Day, y = Observed))+
  geom_point(aes(color = Treatment), size = 4, alpha = 0.75)+
  scale_shape_manual(values = Beverage_shape)+
  scale_color_manual(values = Treatment_color)+
  theme_classic() + labs(x = "Time (Days)", y = "Species Richness")+
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        legend.text = element_text(size = 12))

```

#### Rarefaction curve
```{r}
# 1. Extract the observed richness from iNEXT output
obs_rich_cd <- iNEXT_output_cd$iNextEst$size_based %>%
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Observed") %>%
  # Include sample metadata
  left_join(., metadata_cd, by = "Assemblage") 
  

rarefaction_curve_cd <- 
  # 2. Extract richness calculated from rarefied samples
  iNEXT_output_cd$iNextEst$size_based %>% 
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Rarefaction")%>%
  # Include sample metadata
  left_join(., metadata_cd, by = "Assemblage") %>% 
  # 3. Plot rarefaction curve
  ggplot(aes(x = m/1000, y = qD, color = Treatment, group = Assemblage))+
  scale_color_manual(values = Treatment_color) + 
  geom_line()+ 
  geom_point(data = obs_rich_cd, 
             aes(x = m/1000, y = qD, color = Treatment),
             size = 3)+
  labs(x = "Number of Sequences (x1000)", 
       y = "Species Richness")+ 
  facet_wrap(.~Trial)+
  theme_pubr()

rarefaction_curve_cd

ggpubr::ggarrange(rarefaction_curve_oj, rarefaction_curve_cd,
                  ncol = 1, nrow = 2, 
                  common.legend = TRUE, 
                  label.x = 0.1,
                  label.y = 0.9,
                  legend = "bottom",
                  labels = c("OJ (ITS)", "Cider (ITS)"))
```

```{r}
dunn.test.its.cd <- div_iNEXT_cd_its %>% group_by(Diversity) %>%
  dunn_test(Observed ~ Tech)
dunn.test.its.cd$p.format <- p_format(dunn.test.its.cd$p.adj, accuracy = 0.01)
```

```{r}
alpha_cd_ITS <- div_iNEXT_cd_its %>%
  mutate(Diversity = 
           factor(Diversity, 
                  levels = c("Species richness",
                             "Shannon diversity",
                             "Simpson diversity"))) %>%
  ggplot(aes(x = Tech, y = Observed, color = Treatment, group = Tech))+
  geom_point(alpha = 0.8, size = 3)+
  scale_color_manual(values = Treatment_color)+
  labs(x = NULL, y = "Value")+
  facet_grid(.~Diversity)+
  theme_pubr(legend = "none")+
  theme(axis.text.x = element_text(size = 10, angle = 90))+
  stat_pvalue_manual(dunn.test.its.cd, y.position = c(80,90,100), 
                     label = "p.format")

```

### Diversity boxplot
```{r}
# Diversity in apple cider
obs_rich_cd %>%
  ggplot(aes(x = Shelf_Life_Stage, y = qD, color = Shelf_Life_Stage))+
  geom_boxplot(aes(fill = Shelf_Life_Stage), alpha = 0.5)+
  scale_fill_manual(values = Stage_color)+
  scale_color_manual(values = Stage_color)+
  geom_jitter(size = 2.5, alpha = 0.75)+
  theme_classic()+
  labs(x = "Stage of Shelf Life", 
       y = "Species richness", 
       color = "Stage in Shelf Life",
       fill = "Stage in Shelf Life")+
  facet_grid(.~Thermal)+ stat_kruskal_test()+
  #theme(axis.text.x = element_text(angle = 45))
  theme_pubr()

obs_rich_cd %>%
  ggplot(aes(x = Trial, y = qD, color = Trial))+
  geom_boxplot(aes(fill = Trial), alpha = 0.5)+
  scale_fill_manual(values = Trial_color)+
  scale_color_manual(values = Trial_color)+
  geom_jitter(size = 2.5, alpha = 0.75)+
  theme_classic()+
  labs(x = "Trial", 
       y = "Species richness", 
       color = "Trial",
       fill = "Trial")+
  facet_grid(.~Thermal)+
  stat_kruskal_test()+
  #theme(axis.text.x = element_text(angle = 45))
  theme_pubr()

obs_rich_oj %>%
  filter(Spoilage == "Not spoiled") %>%
  ggplot(aes(x = Trial, y = qD, color = Trial))+
  geom_boxplot(aes(fill = Trial), alpha = 0.5)+
  scale_fill_manual(values = Trial_color)+
  scale_color_manual(values = Trial_color)+
  geom_jitter(size = 2.5, alpha = 0.75)+
  theme_classic()+
  labs(x = "Trial", 
       y = "Species richness", 
       color = "Trial",
       fill = "Trial")+
  facet_grid(Spoilage~Thermal)+
  stat_kruskal_test()+
  #theme(axis.text.x = element_text(angle = 45))
  theme_pubr()


cd_richness <- obs_rich_cd %>%
  ggplot(aes(x = Trial, y = qD))+
  geom_boxplot(aes(color = "black"), 
               outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color,
                     name = "Treatment")+
  labs(y = "Species richness")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test()+
  #theme(axis.text.x = element_text(angle = 45))+
  theme_pubr(legend = "right")
```

### Shannon index for species evenness
```{r}
cd_shannon <-div_iNEXT_cd %>%
  filter(Diversity == "Shannon diversity") %>%
  ggplot(aes(x = Spoilage, y = Observed, fill = Spoilage))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = FALSE)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()+labs(y = "Shannon diversity")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test(label.y = 22)+
  theme_pubr(legend = "right")
```

### Simpson diversity measuring the abundant/dominant taxa
```{r}
 cd_simpson <- div_iNEXT_cd %>%
  filter(Diversity == "Simpson diversity") %>%
  ggplot(aes(x = Spoilage, y = Observed, fill = Spoilage))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = FALSE)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()+labs(y = "Simpson diversity")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test(label.y = 14)+
  theme_pubr(legend = "right")

```


```{r}
alpha_CD_ITS <- div_iNEXT_cd %>%
  mutate(Diversity = 
           factor(Diversity, 
                  levels = c("Species richness",
                             "Shannon diversity",
                             "Simpson diversity"))) %>%
  ggplot(aes(x = Tech, y = Observed))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = FALSE)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  theme_classic()+labs(x = NULL, y = "Value")+
  facet_grid(Trial~Diversity)+
  stat_kruskal_test(label.y = 60)+
  theme_pubr(legend = "bottom")

```


### Combining Shannon and Simpson Diversity Plots
```{r}
ggpubr::ggarrange(cd_richness, cd_shannon, cd_simpson,
                  ncol = 1, nrow = 3, 
                  common.legend = TRUE, 
                  legend = "bottom",
                  labels = c("A", "B", "C"))
ggsave(filename = "Figure/alpha_ITS_cd.png", device = "png", height = 8, width = 8, units = "in")
```

```{r}
div_iNEXT_cd$Diversity = factor(div_iNEXT_cd$Diversity, levels = c("Species richness", "Shannon diversity","Simpson diversity"))

div_iNEXT_cd %>%
  # Filter out 'no treatment' group because sample size too small partly due to failed PCR attempts.
  filter(Thermal != "No Treatment") %>%
  ggplot(aes(Trial, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Thermal~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}")+
  theme_pubr(legend = "right")

div_iNEXT_cd %>%
  filter(Thermal != "No Treatment") %>%
  filter(Treatment != "UV") %>%
  ggplot(aes(Treatment, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Thermal~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}")+
  theme_pubr(legend = "right")

```
##### Session information
```{r}
devtools::session_info()
```