---
title: "Alpha_Diversity_16S"
author: "Yupawadee Galasong"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
### Load packages
```{r}
pacman::p_load(phyloseq, iNEXT, tidyverse, ggpubr, install = FALSE)
```

### Load color and shape preferences
```{r}
source(file = "code/colors.R")

```
### Load input phyloseq object obtained after preprocessing
```{r}
load(file = "Data_16S/full_PERF_physeq.RData")

```
### Biodiversity Analysis: orange juice
```{r}
# Create ASV data frame as an input for iNEXT
iNEXT_input_oj <- 
  PERFect_physeq %>% subset_samples(Beverage == "Orange Juice") %>%
  otu_table() %>% data.frame()

# For reproducibility
set.seed(777)

# Calculate Hill numbers
#q = 0 means total number of taxa ("Effective number of species")
#q = 1 means Number of common taxa (equivalent to exp(Shanon))
#q = 2 means Number of dominant taxa (inverse Simpson)

iNEXT_output_oj <- iNEXT(iNEXT_input_oj, q = c(0,1,2), datatype = "abundance")
iNEXT_output_oj$AsyEst$Assemblage <- gsub('X','',iNEXT_output_oj$AsyEst$Assemblage)
save(iNEXT_output_oj, file = "Data_16S/iNEXT_output_oj.RData")
```

#### Export iNEXT output as a dataframe suitable for plotting 
```{r}
# Creating a data frame containing diversity measures from iNEXT output
# plus metadata
metadata_oj <- 
  PERFect_physeq %>% subset_samples(Beverage == "Orange Juice") %>%
  sample_data() %>% data.frame() %>%
  rownames_to_column(var = "Assemblage") %>%
  # Add another grouping variable - Thermal Vs Non-thermal Vs No Treatment
    mutate(Thermal = case_when(
    (Treatment == "Pasteurization")~"Thermal",
    (Treatment == "HPP")~"Non-thermal",
    (Treatment == "HPP + DMDC + NG")~"Non-thermal",
    (Treatment == "Pasteurization + DMDC + NG")~"Thermal",
    (Treatment == "Pasteurization + SB + PS")~"Thermal",
    (Treatment == "UV")~"Non-thermal",
    (Treatment == "No Treatment")~"No Treatment"))

div_iNEXT_oj <- iNEXT_output_oj$AsyEst %>%
  left_join(., metadata_oj, by = "Assemblage") %>% 
  # Change diversity measures to factor for plotting
  mutate(Diversity = as.factor(Diversity))
```
#### Species richness over shelf-life
```{r}
div_iNEXT_oj %>% 
  filter(Diversity == "Species richness") %>%
  ggplot(aes(x = Day, y = Observed))+
  geom_point(aes(color = Treatment, shape = Trial), size = 3, alpha = 0.75)+
  scale_color_manual(values = Treatment_color)+
  labs(x = "Day", y = "Species Richness")+
  facet_wrap(.~Treatment)+
  guides(color = "none")+
  theme_pubr(legend = "right")
```
#### Rarefaction curve
```{r}
iNEXT_output_oj$iNextEst$size_based$Assemblage <-iNEXT_output_oj$iNextEst$size_based$Assemblage %>% gsub('X','',.)

# 1. Extract the observed richness from iNEXT output
obs_rich_oj <- iNEXT_output_oj$iNextEst$size_based %>%
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Observed") %>%
  # Include sample metadata
  left_join(., metadata_oj, by = "Assemblage") 
  

rarefaction_curve_oj <- 
  # 2. Extract richness calculated from rarefied samples
  iNEXT_output_oj$iNextEst$size_based %>% 
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Rarefaction")%>%
  # Include sample metadata
  left_join(., metadata_oj, by = "Assemblage") %>% 
  # 3. Plot rarefaction curve
  ggplot(aes(x = m/1000, y = qD, 
             color = Treatment, group = Assemblage))+
  scale_color_manual(values = Treatment_color) + 
  geom_line(aes(color = Treatment))+ geom_point(data = obs_rich_oj, 
             aes(x = m/1000, y = qD, 
                 color = Treatment), size = 3)+
  labs(x = "Number of Sequences (x1000)", 
       y = "Species Richness")+
  facet_wrap(.~Trial)+
  theme_pubr()

ggsave(filename = "Figure/rarefaction_curve_16S.png", plot =  rarefaction_curve, device = "png",
       height = 3, width = 6, units = "in")

```

#### Diversity boxplot
```{r}
# Diversity in orange juice
obs_rich_oj %>%
  ggplot(aes(x = Thermal, y = qD, color = Thermal))+
  geom_boxplot(aes(fill = Thermal), alpha = 0.5)+
  scale_fill_manual(values = Thermal_color)+
  scale_color_manual(values = Thermal_color)+
  geom_jitter(size = 2.5, alpha = 0.75)+
  theme_classic()+labs(x = "Treatment", y = "Species richness")+
  facet_grid(.~Shelf_Life_Stage)+
  stat_kruskal_test()+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 45))
```

```{r}
oj_richness <- obs_rich_oj %>%
  ggplot(aes(x = Spoilage, y = qD, fill = Spoilage))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA)+ 
  guides(fill = "none")+
  scale_fill_manual(values = Spoiled_color)+
  guides(fill = FALSE)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  labs(x = NULL, y = "Species richness")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test(label.y = 140)+
  theme_pubr()

```

#### Shannon index for species evenness
```{r}
oj_shannon <- div_iNEXT_oj %>% 
  filter(Diversity == "Shannon diversity") %>%
  ggplot(aes(x = Spoilage, y = Observed, 
             fill = Spoilage))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = "none")+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Spoiled_color)+
  labs(x = NULL, y = "Shannon diversity")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test(label.y = 45)+
  theme_pubr()

```
#### Simpson diversity measuring the abundant/dominant taxa
```{r}
 oj_simpson <- div_iNEXT_oj %>% 
  filter(Diversity == "Simpson diversity") %>% 
  ggplot(aes(x = Spoilage, y = Observed, fill = Spoilage))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = "none")+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Spoiled_color)+
  labs(x = NULL, y = "Simpson diversity")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test(label.y = 25)+
  theme_pubr()
```
#### Combining Shannon and Simpson Diversity Plots

```{r}
ggpubr::ggarrange(oj_richness, oj_shannon, oj_simpson,
                  ncol = 1, nrow = 3, 
                  common.legend = TRUE, 
                  labels = c("A", "B", "C"),
                  legend = "right")
ggsave(filename = "Figure/alpha_16S_oj.png", device = "png", height = 5, width = 7, units = "in")

```
### Biodiversity Analysis: apple cider
```{r}
# Create ASV data frame as an input for iNEXT
iNEXT_input_cd <- 
  PERFect_physeq %>% subset_samples(Beverage == "Cider") %>%
  otu_table() %>% data.frame()

# For reproducibility
set.seed(777)

# Calculate Hill numbers
#q = 0 means total number of taxa ("Effective number of species")
#q = 1 means Number of common taxa (equivalent to exp(Shanon))
#q = 2 means Number of dominant taxa (inverse Simpson)

iNEXT_output_cd <- iNEXT(iNEXT_input_cd, q = c(0,1,2), datatype = "abundance")

iNEXT_output_cd$AsyEst$Assemblage <- gsub('X','',iNEXT_output_cd$AsyEst$Assemblage)
save(iNEXT_output_cd, file = "Data_16S/iNEXT_output_cd.RData")
```

#### Export iNEXT output as a dataframe suitable for plotting 
```{r}
# Creating a data frame containing diversity measures from iNEXT output
# plus metadata
metadata_cd <- 
  PERFect_physeq %>% subset_samples(Beverage == "Cider") %>%
  sample_data() %>% data.frame() %>%
  rownames_to_column(var = "Assemblage") %>%
  # Add another grouping variable - Thermal Vs Non-thermal Vs No Treatment
    mutate(Thermal = case_when(
    (Treatment == "Pasteurization")~"Thermal",
    (Treatment == "HPP")~"Non-thermal",
    (Treatment == "HPP + DMDC + NG")~"Non-thermal",
    (Treatment == "Pasteurization + DMDC + NG")~"Thermal",
    (Treatment == "Pasteurization + SB + PS")~"Thermal",
    (Treatment == "UV")~"Non-thermal",
    (Treatment == "No Treatment")~"No Treatment"))

div_iNEXT_cd <- iNEXT_output_cd$AsyEst %>%
  left_join(., metadata_cd, by = "Assemblage") %>% 
  # Change diversity measures to factor for plotting
  mutate(Diversity = as.factor(Diversity))
```
#### Species richness over shelf-life
```{r}
div_iNEXT_cd %>% 
  filter(Diversity == "Species richness") %>%
  ggplot(aes(x = Day, y = Observed))+
  geom_point(aes(color = Treatment, shape = Trial), size = 3, alpha = 0.75)+
  scale_color_manual(values = Treatment_color)+
  labs(x = "Day", y = "Species Richness")+
  facet_wrap(.~Treatment)+
  guides(color = "none")+
  theme_pubr(legend = "right")
```
#### Rarefaction curve
```{r}
iNEXT_output_cd$iNextEst$size_based$Assemblage <-iNEXT_output_cd$iNextEst$size_based$Assemblage %>% gsub('X','',.)

# 1. Extract the observed richness from iNEXT output
obs_rich_cd <- iNEXT_output_cd$iNextEst$size_based %>%
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Observed") %>%
  # Include sample metadata
  left_join(., metadata_cd, by = "Assemblage") 
  

rarefaction_curve_cd <- 
  # 2. Extract richness calculated from rarefied samples
  iNEXT_output_cd$iNextEst$size_based %>% 
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Rarefaction")%>%
  # Include sample metadata
  left_join(., metadata_cd, by = "Assemblage") %>% 
  # 3. Plot rarefaction curve
  ggplot(aes(x = m/1000, y = qD, 
             color = Treatment, group = Assemblage))+
  scale_color_manual(values = Treatment_color) + 
  geom_line(aes(color = Treatment))+ geom_point(data = obs_rich_cd, 
             aes(x = m/1000, y = qD, 
                 color = Treatment), size = 3)+
  labs(x = "Number of Sequences (x1000)", 
       y = "Species Richness")+
  facet_wrap(.~Trial)+
  theme_pubr()

# Combine oj and cider rarefaction curves
ggarrange(rarefaction_curve_oj, rarefaction_curve_cd, nrow = 2, 
          labels = c("OJ (16S)",
                     "Cider (16S)"),
          label.x = 0.7,
          label.y = 0.9,
          common.legend = TRUE,
          legend = "bottom")
ggsave(filename = "Figure/rarefaction_curve_16S.png", device = "png",
       height = 4, width = 8, units = "in")

```
#### Diversity boxplot
```{r}

# Diversity in cider
obs_rich_cd %>%
  ggplot(aes(x = Thermal, y = qD, color = Thermal))+
  geom_boxplot(aes(fill = Thermal), alpha = 0.5)+
  scale_fill_manual(values = Thermal_color)+
  scale_color_manual(values = Thermal_color)+
  geom_jitter(size = 2.5, alpha = 0.75)+
  theme_classic()+labs(x = "Treatment", y = "Species richness")+
  facet_grid(.~Shelf_Life_Stage)+
  stat_kruskal_test()+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 45))
```

```{r}
cd_richness <- obs_rich_cd %>%
  filter(Thermal != "No Treatment") %>%
  ggplot(aes(x = Trial, y = qD, fill = Trial))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA)+ 
  guides(fill = "none")+
  scale_fill_manual(values = Trial_color)+
  guides(fill = FALSE)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  labs(x = NULL, y = "Species richness")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test()+
  theme_pubr()

```

#### Shannon index for species evenness
```{r}
cd_shannon <- div_iNEXT_cd %>% 
  filter(Thermal != "No Treatment") %>%
  filter(Diversity == "Shannon diversity") %>%
  ggplot(aes(x = Trial, y = Observed, 
             fill = Trial))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = "none")+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Trial_color)+
  labs(x = NULL, y = "Shannon diversity")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test()+
  theme_pubr()

```
#### Simpson diversity measuring the abundant/dominant taxa
```{r}
 cd_simpson <- div_iNEXT_cd %>% 
  filter(Thermal != "No Treatment") %>%
  filter(Diversity == "Simpson diversity") %>% 
  ggplot(aes(x = Trial, y = Observed, fill = Trial))+
  geom_boxplot(alpha = 0.75, outlier.shape = NA) + 
  guides(fill = "none")+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  scale_fill_manual(values = Trial_color)+
  labs(x = NULL, y = "Simpson diversity")+
  facet_grid(.~Thermal)+ 
  stat_kruskal_test()+
  theme_pubr()
```


#### Combining Shannon and Simpson Diversity Plots

```{r}
ggpubr::ggarrange(cd_richness, cd_shannon, cd_simpson,
                  ncol = 2, nrow = 2, 
                  common.legend = TRUE, 
                  labels = c("A", "B", "C"),
                  legend = "bottom")
ggsave(filename = "Figure/alpha_16S_oj.png", device = "png", height = 5, width = 7, units = "in")

```
### Same plot, shorter code
```{r}
#Cider boxplots
div_iNEXT_cd$Diversity = factor(div_iNEXT_cd$Diversity, levels = c("Species richness", "Shannon diversity","Simpson diversity"))

div_iNEXT_cd %>%
  # Filter out 'no treatment' group because sample size too small partly due to failed PCR attempts.
  #filter(Thermal != "No Treatment") %>%
  ggplot(aes(Trial, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Thermal~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}")+
  theme_pubr(legend = "right")

div_iNEXT_cd %>%
  # Filter out 'no treatment' group because sample size too small partly due to failed PCR attempts.
  filter(Thermal != "No Treatment") %>%
  ggplot(aes(Treatment, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Shelf_Life_Stage~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}")+
  theme_pubr(legend = "right")

div_iNEXT_oj %>%
  filter(Thermal != "No Treatment") %>%
  filter(Treatment != "UV") %>%
  ggplot(aes(Trial, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Thermal~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}")+
  theme_pubr(legend = "right")
```
#### Code chunk for playing around
```{r}
# Does adding antimicrobials impact alpha diversity?
# No
div_iNEXT_cd %>% 
  filter(Thermal == "Thermal") %>%
  ggplot(aes(Treatment, y = Observed))+
  geom_boxplot()+
  geom_jitter(aes(color = Treatment, shape = Trial), size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Shelf_Life_Stage~Diversity, scales = "free_y")+
  stat_kruskal_test()

div_iNEXT_oj %>%
  ggplot(aes(Spoilage, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Thermal~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}", label.y = 90)+
  theme_pubr(legend = "right")

div_iNEXT_cd %>%
  ggplot(aes(Spoilage, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Thermal~Diversity, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}", label.y = 90)+
  theme_pubr(legend = "right")
div_iNEXT_cd %>%
  # Filter out 'no treatment' group because sample size too small partly due to failed PCR attempts.
  #filter(Thermal != "No Treatment") %>%
  filter(Shelf_Life_Stage != "Day14") %>%
  ggplot(aes(Shelf_Life_Stage, y = Observed))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = Treatment),size = 2.5)+
  scale_color_manual(values = Treatment_color)+
  facet_grid(Diversity~Thermal, scales = "free_y")+
  stat_kruskal_test(label = "p = {p.format}")+
  labs(x = NULL)+
  theme_pubr(legend = "right")
```
### Session information
```{r}
devtools::session_info()
```