---
title: "Beta_Diversity_ITS"
author: "Yupawadee Galasong"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

### Load packages

```{r}
pacman::p_load(phyloseq, tidyverse, ggpubr, ggforce, tempted, install = FALSE)
```

### Set up color & shape preferences - Metadata variables

Useful palette generator <https://coolors.co/> Make sure it is friendly
to color blind readers.
<https://www.datylon.com/blog/data-visualization-for-colorblind-readers#:~:text=The%20first%20rule%20of%20making,out%20of%20these%20two%20hues.>
Example of how to use color and shape preferences in ggplot
`ggplot(data = final_physeq@sam_data, aes(x = malic, y = fructose))+geom_point(aes(color = Trial, shape = Trial))+scale_color_manual(values = Trial_color)+scale_shape_manual(values = Trial_shape)`

```{r}
# Specify colors
Trial_color <- c("#fcd0a1","#b1b695","#A690A4")
# Specify which trial each color is for
names(Trial_color) <- c(1,2,3)

Trial_shape <- c(15,16,17)
names(Trial_shape) <- c(1,2,3)

Beverage_color <- c("#FFC759","#FF7B9C")
names(Beverage_color) <- c("Orange Juice","Cider")

Beverage_shape <- c(18,19)
names(Beverage_shape) <- c("Orange Juice","Cider")

Treatment_color <- c("#332288", "#882255", "#AA4499", "#CC6677", "#DDCC77", 
                     "#44AA99", "#117733")
names(Treatment_color) <- c("No Treatment", "Pasteurization", 
                           "Pasteurization + SB + PS","Pasteurization + DMDC + NG",
                           "UV","HPP","HPP + DMDC + NG")
Stage_color <- c("#aec5eb", "#3D65A5", "#1F449C")
names(Stage_color) <- c("Beginning", "Day14", "End")

Stage_shape <- c(15,16,17)
names(Stage_shape) <- c("Beginning", "Day14", "End")

Spoiled_color <- c("#FFEAD0", "#F76F8E")
names(Spoiled_color) <- c("Not spoiled", "Spoiled")

Thermal_color <- c("#EB4511","#1C3144","#FFBA08")
names(Thermal_color) <- c("Thermal", "Non-thermal", "No Treatment")
```

## Microbiome data is compositional.

1.  Microbiome Datasets Are Compositional: And This Is Not Optional
    <https://doi.org/10.3389/fmicb.2017.02224>
2.  A Novel Sparse Compositional Technique Reveals Microbial
    Perturbations (robust Aitchison PCA)
    <https://doi.org/10.1128/msystems.00016-19>
3.  Time-Informed Dimensionality Reduction for Longitudinal Microbiome
    Studies <https://doi.org/10.1101/2023.07.26.550749>

### Computational approach for compositional data

Note: the code in the next four chunks won't be run. It is only for
showing the workflow using popular packages. For this project, we will
use a rather novel method implemented in `tempted` package. \####
Imputation of zeros, left-censored and missing data in compositional
data

```{r}
#zCompostionsR::

```

#### Centered log-ration (CLR) transformation

```{r}
#clr <- microbiome::transform(final_physeq, "clr")
#otu_table(clr)[1:5,1:5]
```

#### Aitchison distance

Aitchison distance is basically the Euclidean distances between
CLR-transformed sample abundance vectors.

```{r}
#aitchison <- vegan::vegdist(clr, method = "euclidean")

# Are results the same?
#aitchison2 <- phyloseq::distance(clr, method = "euclidean")
```

#### Variance-based compositional principal component analysis (PCA)

```{r}
## while the method is specified as rda (short for Redundancy Analysis),
## not specifying the 'formula' argument means we are running an unconstrained RDA,
## which is equivalent to running PCA.
# pca_clr <- phyloseq::ordinate(clr, method = "RDA", distance = aitchison2)
```

### TEMPoral TEnsor Decomposition (TEMPTED)

The goal of TEMPTED is to perform dimensionality reduction for
multivariate longitudinal data, with a special attention to longitudinal
microbiome studies.

#### Load the data and modify to the format required by 'tempted' package

```{r}
load("/local/workdir/yg225/JuiceMicrobiome/Data_ITS/metadata.RData")
load("/local/workdir/yg225/JuiceMicrobiome/Data_ITS/final_physeq.RData")

# A data.frame with rows representing samples and matching with data.frame count table and processed table
## Create a unique ID for each (juice, treatment) combination
ID_df <- metadata %>% expand(Beverage, Treatment, Trial) %>% 
  mutate(JuiceID = seq(from = 1, to = nrow(.)))

metatable <- metadata %>% 
  # remove irrelevant columns
  dplyr::select(-c("is.neg","Sample_or_Control")) %>% 
  # Assign JuiceID to each entry
  left_join(., y = ID_df,) %>%
  # Change row names to sample names
  column_to_rownames(var = "Assemblage")



# A data.frame with rows representing samples and matching with data.frame metatable and columns representing microbial features (i.e. OTUs). Each entry is a read count.

counttable <- final_physeq@otu_table %>% t() %>% as.data.frame() 

# Check if rows match and display the number of matches which should equal the number of samples
table(rownames(counttable) == rownames(metatable))

# A variable storing ID and Treatment
metauni <- unique(metatable[,c("JuiceID","Beverage","Treatment","Trial","Thermal")])
```

#### Run TEMPTED for Microbiome Count Data

Following github tutorial
<https://github.com/pixushi/tempted?tab=readme-ov-file>

```{r}
# Remove subject with only 1 time point to avoid error
# An error earlier indicated that JuiceID 25 and 28 must be removed
# Get sample name (ITSXXX) corresponding to JuiceID 25 and 28
to_rm <- metatable %>% dplyr::filter(JuiceID %in% c(25,28)) %>% rownames()

# Remove JuiceID 25 and 28 from metatable
metatable <- metatable %>% dplyr::filter(!(rownames(.)%in% to_rm))

# Remove samples in to_rm from the count table
counttable <- counttable[!(rownames(counttable) %in% to_rm),]
```

### OJ data set

```{r}
counttable_oj <- final_physeq %>% subset_samples(Beverage == "Orange Juice") %>% otu_table() %>% t() %>% as.data.frame()

metatable_oj <- metatable %>% filter(Beverage == "Orange Juice")

res_count_oj <- tempted::tempted_all(counttable_oj,
                                  metatable_oj$Day,
                                  metatable_oj$JuiceID,
                                  #settings used in the tutorial
                                  threshold = 0.95,
                                  transform = "clr",
                                  pseudo = 0.5,
                                  r = 2,
                                  smooth = 1e-5,
                                  pct_ratio = 0.1,
                                  pct_aggregate = 1)
#[output]
#Calculate the 1th Component
#Convergence reached at dif=7.44034992388313e-05, iter=5
#Calculate the 2th Component
#Convergence reached at dif=3.45294506376887e-05, iter=11
```

#### Low-dimensional representation of SUBJECTS

```{r}
# Create a data frame for plotting samples on a 2D space
A_data_oj <- metauni %>% dplyr::filter(!(rownames(.) %in% to_rm)) %>% filter(Beverage == "Orange Juice")

rownames(A_data_oj) <- A_data_oj$JuiceID

# Subject loadings are stored in A_hat (stored in res_count)
A_data_oj <- cbind(res_count_oj$A_hat[rownames(A_data_oj),],A_data_oj)

# Plot of subject loading
p_oj <- A_data_oj %>% 
  ggplot(aes(x = A_data_oj$PC1, y = A_data_oj$PC2))+
  geom_point(aes(shape = Thermal, color = Treatment), 
             size = 4)+ labs(x = "Component 1", y = "Component 2",
                                     title = "Subject loadings")+
  scale_color_manual(values = Treatment_color)
p_oj + geom_mark_ellipse(aes(fill = Trial), expand = unit(5, "mm"))+
scale_fill_manual(values = Trial_color)+
  scale_y_continuous(limits = c(-0.4,0.8))+
  scale_x_continuous(limits = c(-0.4,0.6))+
  theme_classic()
```

Clearly, the beta diversity is driven by Trial, i.e. the source of the
juice.We can also kind of see that samples form thermal VS non-thermal
cluster. The 'no treatment' cluster very much overlap with the
non-thermal cluster. See the plot below.

```{r}
# Plot of subject loading
p_oj2 <- A_data_oj %>% ggplot(aes(x = A_data_oj$PC1, y = A_data_oj$PC2))+
  geom_point(aes(shape = Thermal, color = Treatment), size = 4)+ 
  labs(x = "Component 1", 
       y = "Component 2",
       title = "Subject loadings")+
  scale_color_manual(values = Treatment_color)

p_oj2 + geom_mark_ellipse(aes(fill = Thermal), alpha = 0.15,
                          expand = unit(5, "mm"))+
  scale_fill_manual(values = Thermal_color)+
  scale_y_continuous(limits = c(-0.4,0.6))+
  scale_x_continuous(limits = c(-0.4,0.6))+
  theme_classic()
  

```

#### Temporal loadings

```{r}
# Temporal loadings are stored in Phi_hat in res_count
tempted::plot_time_loading(res_count_oj, r = 2)+
  geom_line(linewidth = 2)+
  labs(title = "Temporal Loadings", x = "Time in shelf-life (Days)")+
  theme_classic()

```

#### Feature (ASV) loadings

```{r}
# Feature loadings are stored in B_hat in res_count
res_count_oj$B_hat %>% as.data.frame() %>%
  ggplot(aes(x = PC1, y = PC2))+
  geom_point() + 
  labs(x = "Component 1", y = "Component 2", title = "Feature loading")+
  theme_classic()

```

#### Log ratio of top features

```{r}
# Log ratios are stored in metafeature_ration in res_count
# For grouping variable, subset the JuiceID, a desired column
tempted::plot_metafeature(res_count_oj$metafeature_ratio, 
                          metauni[,c(1,5)],nrow = 2) + 
  geom_line(color = "black")+
  xlab("Days in shelf-life")+theme_classic()+
  scale_color_manual(values = Thermal_color, name = "Treatment")+
  scale_fill_manual(values = Thermal_color, name = "Treatment")
```

#### Subject trajectories

```{r}
# Subject trajectories are stored in metafeature_aggregate in res_count
tempted::plot_metafeature(res_count_oj$metafeature_aggregate, metauni[,c(1,5)],nrow = 2)+
  xlab("Days in shelf-life")+theme_classic()+
  scale_fill_manual(values = Thermal_color, name = "Treatment")+
  scale_color_manual(values = Thermal_color, name = "Treatment")

```

#### Low-dimensional representation of SAMPLES

```{r}
tab_feat_obs_oj <- res_count_oj$metafeature_aggregate %>%
  as.data.frame()%>%
  rename(JuiceID = subID) %>%
  mutate(JuiceID = as.integer(JuiceID)) %>%
  left_join(.,metauni)

reshape_feat_obs_oj <- reshape(tab_feat_obs_oj,
                            idvar = c("JuiceID", "timepoint"),
                            v.names = c("value"),
                            timevar = "PC",
                            direction = "wide")
colnames(reshape_feat_obs_oj) <- sub(".*value[.]","", colnames(reshape_feat_obs_oj))
metatable_oj$timepoint <- metatable_oj$Day
reshape_feat_obs_oj <- reshape_feat_obs_oj %>% 
  left_join(., y = metatable_oj, by = c("Treatment", "Trial", "timepoint")) 


reshape_feat_obs_oj %>% 
  ggplot(aes(x = PC1, y = PC2))+
  geom_point(aes(color = timepoint, shape = Thermal.x),size = 4)+ gradient_color("RdYlBu")+
  labs(x = "Component 1", y = "Component 2", color = "Days")+
  geom_mark_ellipse(aes(fill = Thermal.x), alpha = 0.15, expand = unit(0, "mm"))+ 
  scale_fill_manual(values = Thermal_color)+
  theme_classic()
  #theme(plot.margin = margin(2, 2, 2, 2, "cm"))+
  #scale_x_continuous(limits = c(-10,22))+
  #scale_y_continuous(limits = c(-22,24))

```

This plot below shows that we cannot distinguish spoiled vs not spoiled

```{r}
reshape_feat_obs_oj %>% 
  filter(Shelf_Life_Stage == "End") %>%
  ggplot(aes(x = PC1, y = PC2))+
  geom_point(aes(color = Spoilage, shape = Thermal.x),size = 4)+ 
  scale_color_manual(values = Spoiled_color)+
  labs(x = "Component 1", y = "Component 2", color = "Spoiled?", shape = "Treatment")+
  geom_mark_ellipse(aes(fill = Spoilage), 
                    alpha = 0.15, expand = unit(1, "mm"))+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()+
  theme(plot.margin = margin(2, 2, 2, 2, "cm"))+
  scale_x_continuous(limits = c(-10,24))+
  scale_y_continuous(limits = c(-22,24))
```

Trajectory of Top features Component 2

```{r}
proportion_table_oj <- counttable_oj/rowSums(counttable_oj)

top_ASVs <- res_count_oj$toppct_ratio %>% as.data.frame() %>% filter(PC2 == TRUE) %>% rownames()

final_physeq@tax_table[top_ASVs][,c("Genus","Species")]
top_ASVs_PC2_labs <- c("Penicillium sp. 1",
                       "Penicillium sp. 2",
                       "Colletotrichum sp.",
                       "Lasiodiplodia sp.",
                       "Alternaria sp. 1",
                       "Diaporthe sp.",
                       "Botrytis sp.",
                       "Fusarium sp.",
                       "Eremothecium sp.",
                       "Unclassified Didymellaceae",
                       "Alternaria sp. 1",
                       "Alternaria sp. 2",
                       "Meyerozyma sp",
                       "Diutina sp.")
names(top_ASVs_PC2_labs) <- top_ASVs
tab_sel_feat_oj <- cbind(proportion_table_oj[,top_ASVs], metatable_oj)
tab_sel_feat_oj <- reshape(tab_sel_feat_oj, 
                        varying=list(top_ASVs), 
                        times=top_ASVs, 
                        timevar="feature", 
                        v.names="RA", 
                        ids=rownames(tab_sel_feat_oj),
                        direction="long")
p_topfeat_oj <- ggplot(data=tab_sel_feat_oj) + 
  geom_point(aes(x=Day, y=RA, color=Trial, shape = Thermal), 
             size = 2) +
  facet_wrap(vars(feature),  labeller = labeller(feature = top_ASVs_PC2_labs))+ scale_color_manual(values = Trial_color) + 
  labs(x="Day of Shelf-Life", y="Relative Abundance")+  
  coord_trans(y="sqrt")+theme_pubr()
# summary plot
p_topfeat_summary_oj <- plot_feature_summary(proportion_table_oj[,top_ASVs], 
                     metatable_oj$Day, 
                     metatable_oj$Spoilage, 
                     bws=30, nrow = 4) + 
  labs(x="Day of Shelf-Life", y="Relative Abundance")+  
  coord_trans(y="sqrt")

```

Component 1

```{r}
top_ASVs_PC1 <- res_count_oj$toppct_ratio %>% as.data.frame() %>% filter(PC1 == TRUE) %>% rownames()
final_physeq@tax_table[top_ASVs_PC1][,c("Genus","Species")]
top_ASVs_PC1_labs <- c("Penicillium sp. 1",
                       "Saccharomyces sp.",
                       "Pichia sp. 1",
                       "Pichia sp. 2",
                       "Clavispora sp.",
                       "Pichia sp. 3",
                       "Penicillium sp. 2",
                       "Colletotrichum sp.",
                       "Pichia sp. 4",
                       "Starmerella sp. 1",
                       "Pichia sp. 5",
                       "Starmerella sp. 2",
                       "Candida sp.",
                       "Starmera sp.")
names(top_ASVs_PC1_labs) <- top_ASVs_PC1
tab_sel_feat_oj_PC1 <- cbind(proportion_table_oj[,top_ASVs_PC1], metatable_oj)
tab_sel_feat_oj_PC1 <- reshape(tab_sel_feat_oj_PC1, 
                        varying=list(top_ASVs_PC1), 
                        times=top_ASVs_PC1, 
                        timevar="feature", 
                        v.names="RA", 
                        ids=rownames(tab_sel_feat_oj_PC1),
                        direction="long")
p_topfeat_oj_PC1 <- ggplot(data=tab_sel_feat_oj_PC1) + 
  geom_point(aes(x=Day, y=RA, color=APDA, shape = Thermal), 
             size = 2) +
  facet_wrap(vars(feature), labeller = labeller(feature = top_ASVs_PC1_labs))+ gradient_color("OrRd")+
  labs(x="Day of Shelf-Life", y="Relative Abundance")+  
  coord_trans(y="sqrt")+theme_pubr()

# summary plot

p_topfeat_summary_oj_PC1 <- plot_feature_summary(proportion_table_oj[,top_ASVs_PC1], 
                     metatable_oj$Day, 
                     metatable_oj$Thermal, 
                     bws=30, nrow = 4) + 
  labs(x="Day of Shelf-Life", y="Relative Abundance")
str(p_topfeat_summary_oj_PC1)

```
### Cider data set

```{r}
# Remove subject with only 1 time point to avoid error
# An error earlier indicated that JuiceID 25 and 28 must be removed
# Get sample name (ITSXXX) corresponding to JuiceID 25 and 28
to_rm <- metatable %>% dplyr::filter(JuiceID %in% c(25,28)) %>% rownames()

# Remove JuiceID 25 and 28 from metatable
metatable <- metatable %>% dplyr::filter(!(rownames(.)%in% to_rm))

# Remove samples in to_rm from the count table
counttable <- counttable[!(rownames(counttable) %in% to_rm),]
```

```{r}
counttable_cd <- final_physeq %>% subset_samples(Beverage == "Cider") %>% otu_table() %>% t() %>% as.data.frame() %>% dplyr::filter(!(rownames(.)%in% to_rm))

metatable_cd <- metatable %>% filter(Beverage == "Cider")
res_count_cd <- tempted::tempted_all(counttable_cd,
                                  metatable_cd$Day,
                                  metatable_cd$JuiceID,
                                  #settings used in the tutorial
                                  threshold = 0.95,
                                  transform = "clr",
                                  pseudo = 0.5,
                                  r = 2,
                                  smooth = 1e-5,
                                  pct_ratio = 0.1,
                                  pct_aggregate = 1)
#[output]
#Calculate the 1th Component
#Convergence reached at dif=4.6052402294862e-05, iter=5
#Calculate the 2th Component
#Convergence reached at dif=4.04833717349942e-05, iter=4
```

#### Low-dimensional representation of SUBJECTS

```{r}
# Create a data frame for plotting samples on a 2D space
A_data_cd <- metauni %>% dplyr::filter(!(rownames(.) %in% to_rm)) %>% filter(Beverage == "Cider")

rownames(A_data_cd) <- A_data_cd$JuiceID

# Subject loadings are stored in A_hat (stored in res_count)
A_data_cd <- cbind(res_count_cd$A_hat[rownames(A_data_cd),],A_data_cd)

# Plot of subject loading
p_cd <- A_data_cd %>% 
  ggplot(aes(x = A_data_cd$PC1, y = A_data_cd$PC2))+
  geom_point(aes(shape = Thermal, color = Treatment), 
             size = 4)+ labs(x = "Component 1", y = "Component 2",
                                     title = "Subject loadings")+
  scale_color_manual(values = Treatment_color)
p_cd + geom_mark_ellipse(aes(fill = Trial), expand = unit(3, "mm"))+
scale_fill_manual(values = Trial_color)+
  scale_y_continuous(limits = c(-0.4,0.8))+
  scale_x_continuous(limits = c(-0.4,0.6))+
  theme_classic()
```

Clearly, the beta diversity is driven by Trial, i.e. the source of the
juice.Unlike in orange juice, cider samples don't seem to form thermal Vs non-thermal clusters.

```{r}
# Plot of subject loading
p_cd2 <- A_data_cd %>% ggplot(aes(x = A_data_cd$PC1, y = A_data_cd$PC2))+
  geom_point(aes(shape = Thermal, color = Treatment), size = 4)+ 
  labs(x = "Component 1", 
       y = "Component 2",
       title = "Subject loadings")+
  scale_color_manual(values = Treatment_color)

p_cd2 + geom_mark_ellipse(aes(fill = Thermal), alpha = 0.15,
                          expand = unit(0, "mm"))+
  scale_fill_manual(values = Thermal_color)+
  scale_y_continuous(limits = c(-0.4,0.8))+
  scale_x_continuous(limits = c(-0.6,0.6))+
  theme_classic()
  

```

#### Temporal loadings

```{r}
# Temporal loadings are stored in Phi_hat in res_count
tempted::plot_time_loading(res_count_cd, r = 2)+
  geom_line(linewidth = 2)+
  labs(title = "Temporal Loadings", x = "Time in shelf-life (Days)")+
  theme_classic()

```

#### Feature (ASV) loadings

```{r}
# Feature loadings are stored in B_hat in res_count
res_count_cd$B_hat %>% as.data.frame() %>%
  ggplot(aes(x = PC1, y = PC2))+
  geom_point() + 
  labs(x = "Component 1", y = "Component 2", title = "Feature loading")+
  theme_classic()

```

#### Log ratio of top features

```{r}
# Log ratios are stored in metafeature_ration in res_count
# For grouping variable, subset the JuiceID, a desired column
tempted::plot_metafeature(res_count_cd$metafeature_ratio, 
                          metauni[,c(1,5)],nrow = 2) + 
  geom_line(color = "black")+
  xlab("Days in shelf-life")+theme_classic()+
  scale_color_manual(values = Thermal_color, name = "Treatment")+
  scale_fill_manual(values = Thermal_color, name = "Treatment")
```

#### Subject trajectories

```{r}
# Subject trajectories are stored in metafeature_aggregate in res_count
tempted::plot_metafeature(res_count_cd$metafeature_aggregate, metauni[,c(1,5)],nrow = 2)+
  xlab("Days in shelf-life")+theme_classic()+
  scale_fill_manual(values = Thermal_color, name = "Treatment")+
  scale_color_manual(values = Thermal_color, name = "Treatment")

```

#### Low-dimensional representation of SAMPLES

```{r}
tab_feat_obs_cd <- res_count_cd$metafeature_aggregate %>%
  as.data.frame()%>%
  rename(JuiceID = subID) %>%
  mutate(JuiceID = as.integer(JuiceID)) %>%
  left_join(.,metauni)

reshape_feat_obs_cd <- reshape(tab_feat_obs_cd,
                            idvar = c("JuiceID", "timepoint"),
                            v.names = c("value"),
                            timevar = "PC",
                            direction = "wide")
colnames(reshape_feat_obs_cd) <- sub(".*value[.]","", colnames(reshape_feat_obs_cd))
metatable_cd$timepoint <- metatable_cd$Day
reshape_feat_obs_cd <- reshape_feat_obs_cd %>% 
  left_join(., y = metatable_cd, by = c("Treatment", "Trial", "timepoint")) 


reshape_feat_obs_cd %>% 
  ggplot(aes(x = PC1, y = PC2))+
  geom_point(aes(color = timepoint, shape = Thermal.x),size = 4)+ gradient_color("RdYlBu")+
  labs(x = "Component 1", y = "Component 2", color = "Days")+
  geom_mark_ellipse(aes(fill = Thermal.x), alpha = 0.15, expand = unit(0, "mm"))+ 
  scale_fill_manual(values = Thermal_color)+
  theme_classic()
  #theme(plot.margin = margin(2, 2, 2, 2, "cm"))+
  #scale_x_continuous(limits = c(-10,22))+
  #scale_y_continuous(limits = c(-22,24))

```

This plot below shows that we cannot distinguish spoiled vs not spoiled

```{r}
reshape_feat_obs_cd %>% 
  filter(Shelf_Life_Stage == "End") %>%
  ggplot(aes(x = PC1, y = PC2))+
  geom_point(aes(color = Spoilage, shape = Thermal.x),size = 4)+ 
  scale_color_manual(values = Spoiled_color)+
  labs(x = "Component 1", y = "Component 2", color = "Spoiled?", shape = "Treatment")+
  geom_mark_ellipse(aes(fill = Spoilage), 
                    alpha = 0.15, expand = unit(1, "mm"))+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()+
  theme(plot.margin = margin(2, 2, 2, 2, "cm"))+
  scale_x_continuous(limits = c(-22,22))+
  scale_y_continuous(limits = c(-10,22))
```

Trajectory of Top features Component 2

```{r}
proportion_table_cd <- counttable_cd/rowSums(counttable_cd)

top_ASVs <- res_count_cd$toppct_ratio %>% as.data.frame() %>% filter(PC2 == TRUE) %>% rownames()

final_physeq@tax_table[top_ASVs][,c("Family","Genus","Species")]
top_ASVs_PC2_labs <- c("Unclassified \n Saccharomycetales 1",
                       "Cladosporium sp.",
                       "Candida sp.",
                       "Pichia sp.",
                       "Unclassified \n Saccharomycetales 2",
                       "Kurtzmaniella sp.",
                       "Chaetocapnodium sp. 1",
                       "Didymella sp. 1",
                       "Bulleromyces sp.",
                       "Hanseniaspora sp.",
                       "Aureobasidium sp.",
                       "Unclassified \n Didymellaceae",
                       "Chaetocapnodium sp. 2",
                       "Didymella sp. 2",
                       "Saccharomycopsis sp.",
                       "Zygosaccharomyces sp.",
                       "Oidiodendron sp.",
                       "Melanchlenus sp.")
names(top_ASVs_PC2_labs) <- top_ASVs
tab_sel_feat_cd <- cbind(proportion_table_cd[,top_ASVs], metatable_cd)
tab_sel_feat_cd <- reshape(tab_sel_feat_cd, 
                        varying=list(top_ASVs), 
                        times=top_ASVs, 
                        timevar="feature", 
                        v.names="RA", 
                        ids=rownames(tab_sel_feat_cd),
                        direction="long")
p_topfeat_cd <- ggplot(data=tab_sel_feat_cd) + 
  geom_point(aes(x=Day, y=RA, color=Thermal, shape = Trial), alpha = 0.75,
             size = 3) +
  facet_wrap(vars(feature),  labeller = labeller(feature = top_ASVs_PC2_labs), nrow = 4)+ scale_color_manual(values = Thermal_color) + 
  labs(x="Day of Shelf-Life", y="Relative Abundance")+  
  coord_trans(y="sqrt")+theme_pubr()
# summary plot
p_topfeat_summary_cd <- plot_feature_summary(proportion_table_cd[,top_ASVs], 
                     metatable_cd$Day, 
                     metatable_cd$Spoilage, 
                     bws=30, nrow = 4) + 
  labs(x="Day of Shelf-Life", y="Relative Abundance")

```

Component 1

```{r}
top_ASVs_PC1 <- res_count_cd$toppct_ratio %>% as.data.frame() %>% filter(PC1 == TRUE) %>% rownames()
final_physeq@tax_table[top_ASVs_PC1][,c("Family","Genus","Species")]
top_ASVs_PC1_labs <- c("Mrakia sp. 1",
                       "Penicillium sp.",
                       "Mrakia sp. 2",
                       "Mrakia sp. 3",
                       "Cystofilobasidium sp. 1",
                       "Mrakia sp. 4",
                       "Mrakia sp. 5",
                       "Pichia sp. 1",
                       "Cephaloascus sp.",
                       "Cystofilobasidium sp. 2",
                       "Pichia sp. 2",
                       "Metschnikowia sp.",
                       "Starmerella sp.",
                       "Saturnispora sp.",
                       "Unclassified \n Mrakiaceae",
                       "Cystofilobasidium sp. 3",
                       "Mrakia sp. 6",
                       "Rhizophaera sp.")
names(top_ASVs_PC1_labs) <- top_ASVs_PC1
tab_sel_feat_cd_PC1 <- cbind(proportion_table_cd[,top_ASVs_PC1], metatable_cd)
tab_sel_feat_cd_PC1 <- reshape(tab_sel_feat_cd_PC1, 
                        varying=list(top_ASVs_PC1), 
                        times=top_ASVs_PC1, 
                        timevar="feature", 
                        v.names="RA", 
                        ids=rownames(tab_sel_feat_cd_PC1),
                        direction="long")
p_topfeat_cd_PC1 <- ggplot(data=tab_sel_feat_cd_PC1) + 
  geom_point(aes(x=Day, y=RA, color=Thermal, shape = Spoilage), alpha = 0.75, size = 3) +
  facet_wrap(vars(feature), labeller = labeller(feature = top_ASVs_PC1_labs))+ #gradient_color("OrRd")+
  scale_color_manual(values = Thermal_color)+
  labs(x="Day of Shelf-Life", y="Relative Abundance")+  
  coord_trans(y="sqrt")+theme_pubr()

# summary plot

p_topfeat_summary_cd_PC1 <- plot_feature_summary(proportion_table_cd[,top_ASVs_PC1], 
                     metatable_cd$Day, 
                     metatable_cd$Thermal, 
                     bws=30, nrow = 4) + 
  labs(x="Day of Shelf-Life", y="Relative Abundance")+
  guides(color = guide_legend(title = "Treatment"),
         fill = guide_legend(title = "Treatment")) +
  scale_fill_manual(values = Thermal_color)+
  scale_color_manual(values = Thermal_color)+
  facet_grid(labeller = top_ASVs_PC1_labs)+
  facet_wrap(vars(feature_all), scales = "free_y", labeller = labeller(feature_all = top_ASVs_PC1_labs))+
  theme_pubr()

```