---
title: "Beta_Diversity_ITS"
author: "Yupawadee Galasong"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
### Load packages
```{r}
library(zCompositions)
library(phyloseq)
library(tidyverse)
library(ggpubr)
library(vegan)
library(microbiome)
library(tempted)
library(ggforce)
```

### Set up color & shape preferences - Metadata variables
Useful palette generator <https://coolors.co/>
Make sure it is friendly to color blind readers.
<https://www.datylon.com/blog/data-visualization-for-colorblind-readers#:~:text=The%20first%20rule%20of%20making,out%20of%20these%20two%20hues.>
Example of how to use color and shape preferences in ggplot
`ggplot(data = final_physeq@sam_data, aes(x = malic, y = fructose))+geom_point(aes(color = Trial, shape = Trial))+scale_color_manual(values = Trial_color)+scale_shape_manual(values = Trial_shape)`
```{r}
# Specify colors
Trial_color <- c("#fcd0a1","#b1b695","#A690A4")
# Specify which trial each color is for
names(Trial_color) <- c(1,2,3)

Trial_shape <- c(15,16,17)
names(Trial_shape) <- c(1,2,3)

Beverage_color <- c("#FFC759","#FF7B9C")
names(Beverage_color) <- c("Orange Juice","Cider")

Beverage_shape <- c(18,19)
names(Beverage_shape) <- c("Orange Juice","Cider")

Treatment_color <- c("#332288", "#882255", "#AA4499", "#CC6677", "#DDCC77", 
                     "#44AA99", "#117733")
names(Treatment_color) <- c("No Treatment", "Pasteurization", 
                           "Pasteurization + SB + PS","Pasteurization + DMDC + NG",
                           "UV","HPP","HPP + DMDC + NG")
Stage_color <- c("#7CA1CC", "#3D65A5", "#1F449C")
names(Stage_color) <- c("Beginning", "Day14", "End")

Stage_shape <- c(15,16,17)
names(Stage_shape) <- c("Beginning", "Day14", "End")

Spoiled_color <- c("#FFEAD0", "#F76F8E")
names(Spoiled_color) <- c("Not spoiled", "Spoiled")

Thermal_color <- c("#EB4511","#1C3144","#FFBA08")
names(Thermal_color) <- c("Thermal", "Non-thermal", "No Treatment")
```
### Set up color & shape preferences - Taxanomical ranks
```{r}


```

## Microbiome data is compositional.
1. Microbiome Datasets Are Compositional: And This Is Not Optional
<https://doi.org/10.3389/fmicb.2017.02224>
2. A Novel Sparse Compositional Technique Reveals Microbial Perturbations
(robust Aitchison PCA)
<https://doi.org/10.1128/msystems.00016-19>
3. Time-Informed Dimensionality Reduction for Longitudinal Microbiome Studies
<https://doi.org/10.1101/2023.07.26.550749>

### Computational approach for compositional data
Note: the code in the next four chunks won't be run. It is only for showing the workflow
using popular packages. For this project, we will use a rather novel method implemented in
`tempted` package.
####  Imputation of zeros, left-censored and missing data in compositional data 
```{r}
#zCompostionsR::

```

#### Centered log-ration (CLR) transformation
```{r}
#clr <- microbiome::transform(final_physeq, "clr")
#otu_table(clr)[1:5,1:5]
```

#### Aitchison distance
Aitchison distance is basically the Euclidean distances between CLR-transformed 
sample abundance vectors.
```{r}
#aitchison <- vegan::vegdist(clr, method = "euclidean")

# Are results the same?
#aitchison2 <- phyloseq::distance(clr, method = "euclidean")
```

#### Variance-based compositional principal component analysis (PCA) 
```{r}
## while the method is specified as rda (short for Redundancy Analysis),
## not specifying the 'formula' argument means we are running an unconstrained RDA,
## which is equivalent to running PCA.
# pca_clr <- phyloseq::ordinate(clr, method = "RDA", distance = aitchison2)
```

### TEMPoral TEnsor Decomposition (TEMPTED)
The goal of TEMPTED is to perform dimensionality reduction for multivariate longitudinal data, with a special attention to longitudinal microbiome studies.

#### Load the data and modify to the format required by 'tempted' package
```{r}
load("/local/workdir/yg225/JuiceMicrobiome/Data_ITS/metadata.RData")
load("/local/workdir/yg225/JuiceMicrobiome/Data_ITS/final_physeq.RData")

# A data.frame with rows representing samples and matching with data.frame count table and processed table
## Create a unique ID for each (juice, treatment) combination
ID_df <- metadata %>% expand(Beverage, Treatment, Trial) %>% 
  mutate(JuiceID = seq(from = 1, to = nrow(.)))

metatable <- metadata %>% 
  # remove irrelevant columns
  dplyr::select(-c("is.neg","Sample_or_Control")) %>% 
  # Assign JuiceID to each entry
  left_join(., y = ID_df,) %>%
  # Change row names to sample names
  column_to_rownames(var = "Assemblage")



# A data.frame with rows representing samples and matching with data.frame metatable and columns representing microbial features (i.e. OTUs). Each entry is a read count.

counttable <- final_physeq@otu_table %>% t() %>% as.data.frame() 

# Check if rows match and display the number of matches which should equal the number of samples
table(rownames(counttable) == rownames(metatable))

# A variable storing ID and Treatment
metauni <- unique(metatable[,c("JuiceID","Beverage","Treatment","Trial","Thermal")])
```

#### Run TEMPTED for Microbiome Count Data
Following github tutorial <https://github.com/pixushi/tempted?tab=readme-ov-file>
```{r}
# Remove subject with only 1 time point to avoid error
# An error earlier indicated that JuiceID 25 and 28 must be removed
# Get sample name (ITSXXX) corresponding to JuiceID 25 and 28
to_rm <- metatable %>% dplyr::filter(JuiceID %in% c(25,28)) %>% rownames()

# Remove JuiceID 25 and 28 from metatable
metatable <- metatable %>% dplyr::filter(!(rownames(.)%in% to_rm))

# Remove samples in to_rm from the count table
counttable <- counttable[!(rownames(counttable) %in% to_rm),]
```
#### OJ data set
```{r}
counttable_oj <- final_physeq %>% subset_samples(Beverage == "Orange Juice") %>% otu_table() %>% t() %>% as.data.frame()

metatable_oj <- metatable %>% filter(Beverage == "Orange Juice")

res_count_oj <- tempted::tempted_all(counttable_oj,
                                  metatable_oj$Day,
                                  metatable_oj$JuiceID,
                                  #settings used in the tutorial
                                  threshold = 0.95,
                                  transform = "clr",
                                  pseudo = 0.5,
                                  r = 2,
                                  smooth = 1e-5,
                                  pct_ratio = 0.1,
                                  pct_aggregate = 1)
#[output]
#Calculate the 1th Component
#Convergence reached at dif=7.44034992388313e-05, iter=5
#Calculate the 2th Component
#Convergence reached at dif=3.45294506376887e-05, iter=11
```
#### Low-dimensional representation of SUBJECTS
```{r}
# Create a data frame for plotting samples on a 2D space
A_data_oj <- metauni %>% dplyr::filter(!(rownames(.) %in% to_rm)) %>% filter(Beverage == "Orange Juice")

rownames(A_data_oj) <- A_data_oj$JuiceID

# Subject loadings are stored in A_hat (stored in res_count)
A_data_oj <- cbind(res_count_oj$A_hat[rownames(A_data_oj),],A_data_oj)

# Plot of subject loading
p_oj <- A_data_oj %>% 
  ggplot(aes(x = A_data_oj$PC1, y = A_data_oj$PC2))+
  geom_point(aes(shape = Thermal, color = Thermal), 
             size = 4)+ labs(x = "Component 1", y = "Component 2",
                                     title = "Subject loadings")+
  scale_color_manual(values = Thermal_color)
p_oj + geom_mark_ellipse(aes(fill = Thermal), expand = unit(5, "mm"))+
scale_fill_manual(values = Thermal_color)+
  scale_y_continuous(limits = c(-0.4,0.6))+
  scale_x_continuous(limits = c(-0.4,0.6))+
  theme_classic()
```

Clearly, the beta diversity is driven by Trial, i.e. the source of the juice.We can also kind of see that samples form thermal VS non-thermal cluster. The 'no treatment' cluster very much overlap with the non-thermal cluster.
See the plot below.
```{r}
# Plot of subject loading
p_oj2 <- A_data_oj %>% ggplot(aes(x = A_data_oj$PC1, y = A_data_oj$PC2))+
  geom_point(aes(shape = Thermal, color = Treatment), size = 4)+ 
  labs(x = "Component 1", 
       y = "Component 2",
       title = "Subject loadings")+
  scale_color_manual(values = Treatment_color)

p_oj2 + geom_mark_ellipse(aes(fill = Thermal), alpha = 0.15,
                          expand = unit(5, "mm"))+
  scale_fill_manual(values = Thermal_color)+
  scale_y_continuous(limits = c(-0.4,0.6))+
  scale_x_continuous(limits = c(-0.4,0.6))+
  theme_classic()
  

```

#### Temporal loadings
```{r}
# Temporal loadings are stored in Phi_hat in res_count
tempted::plot_time_loading(res_count_oj, r = 2)+
  geom_line(linewidth = 2)+
  labs(title = "Temporal Loadings", x = "Time in shelf-life (Days)")+
  theme_classic()

```

#### Feature (ASV) loadings
```{r}
# Feature loadings are stored in B_hat in res_count
res_count_oj$B_hat %>% as.data.frame() %>%
  ggplot(aes(x = PC1, y = PC2))+
  geom_point() + 
  labs(x = "Component 1", y = "Component 2", title = "Feature loading")+
  theme_classic()

```

#### Log ratio of top features
```{r}
# Log ratios are stored in metafeature_ration in res_count
# For grouping variable, subset the JuiceID, a desired column
tempted::plot_metafeature(res_count_oj$metafeature_ratio, 
                          metauni[,c(1,5)],nrow = 2) + 
  geom_line(color = "black")+
  xlab("Days in shelf-life")+theme_classic()+
  scale_color_manual(values = Thermal_color, name = "Treatment")+
  scale_fill_manual(values = Thermal_color, name = "Treatment")
```
#### Subject trajectories
```{r}
# Subject trajectories are stored in metafeature_aggregate in res_count
tempted::plot_metafeature(res_count_oj$metafeature_aggregate, metauni[,c(1,5)],nrow = 2)+
  xlab("Days in shelf-life")+theme_classic()+
  scale_fill_manual(values = Thermal_color, name = "Treatment")+
  scale_color_manual(values = Thermal_color, name = "Treatment")

```
#### Low-dimensional representation of SAMPLES
```{r}
tab_feat_obs_oj <- res_count_oj$metafeature_aggregate %>%
  as.data.frame()%>%
  rename(JuiceID = subID) %>%
  mutate(JuiceID = as.integer(JuiceID)) %>%
  left_join(.,metauni)

reshape_feat_obs_oj <- reshape(tab_feat_obs_oj,
                            idvar = c("JuiceID", "timepoint"),
                            v.names = c("value"),
                            timevar = "PC",
                            direction = "wide")
colnames(reshape_feat_obs_oj) <- sub(".*value[.]","", colnames(reshape_feat_obs_oj))
metatable_oj$timepoint <- metatable_oj$Day
reshape_feat_obs_oj <- reshape_feat_obs_oj %>% 
  left_join(., y = metatable_oj, by = c("Treatment", "Trial", "timepoint")) 


reshape_feat_obs_oj %>% 
  ggplot(aes(x = PC1, y = PC2))+
  geom_point(aes(color = timepoint, shape = Thermal.x),size = 4)+ gradient_color("RdYlBu")+
  labs(x = "Component 1", y = "Component 2", color = "Days")+
  geom_mark_ellipse(aes(fill = Thermal.x), alpha = 0.15, expand = unit(0, "mm"))+ 
  scale_fill_manual(values = Thermal_color)+
  theme_classic()
  #theme(plot.margin = margin(2, 2, 2, 2, "cm"))+
  #scale_x_continuous(limits = c(-10,22))+
  #scale_y_continuous(limits = c(-22,24))

```
This plot below shows that we cannot distinguish spoiled vs not spoiled
```{r}
reshape_feat_obs_oj %>% 
  filter(Shelf_Life_Stage == "End") %>%
  ggplot(aes(x = PC1, y = PC2))+
  geom_point(aes(color = Spoilage, shape = Thermal.x),size = 4)+ 
  #gradient_color("RdYlBu")+
  labs(x = "Component 1", y = "Component 2", color = "APDA")+
  geom_mark_ellipse(aes(fill = Spoilage), 
                    alpha = 0.15, expand = unit(1, "mm"))+
  scale_fill_manual(values = Spoiled_color)+
  theme_classic()
#theme(plot.margin = margin(2, 2, 2, 2, "cm"))+
  #scale_x_continuous(limits = c(-10,22))+
  #scale_y_continuous(limits = c(-22,24))
```

Trajectory of Top features
