---
title: "Compositional_Barplots"
author: "Yupawadee Galasong"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Load required packages
```{r}
pacman::p_load(phyloseq, tidyverse, microViz, ggpubr, install = FALSE)

```

### Load 16S and ITS phyloseq objects
```{r}
load(file = "Data_16S/full_PERF_physeq.RData")

#Rename the object because 16S and ITS physeq have the same name coming out of the same pre-processing workflow.
physeq_16S <- PERFect_physeq

load(file = "Data_ITS/PERFect_physeq.RData")
physeq_ITS <- PERFect_physeq

#Delete the object to keep just two phyloseq objects
rm(PERFect_physeq)
```

### Prepare data frames for compositional plots
Tutorial: <https://rpubs.com/lgschaerer/1006964>
#### 1. Convert phyloseq objects to data frames
```{r}
genus_16S <- physeq_16S %>%
  # Agglomerate taxa to "Genus" level
  tax_glom(taxrank = "Genus") %>%  
  # Transform to relative abundance
  transform_sample_counts(function(x) {x/sum(x)} ) %>%   
  # Melt to long format using phyloseq's melt function
  psmelt()                                               
head(genus_16S)
```
With the ITS taxonomic data, we need to remove the "p__" from "p__(Phylum name)".
```{r}
# Replace "p__" with blank
physeq_ITS@tax_table <- gsub("[[:alnum:]]__","",physeq_ITS@tax_table)

genus_ITS <- physeq_ITS %>%
  # Agglomerate taxa to "Genus" level
  tax_glom(taxrank = "Genus") %>%  
  # Transform to relative abundance
  transform_sample_counts(function(x) {x/sum(x)} ) %>%   
  # Melt to long format using phyloseq's melt function
  psmelt()                                               
head(genus_ITS)

```
#### 2. Filter & modify data frames for plotting

```{r}
# OJ 16S data
filt_16S_oj <- genus_16S %>%
  filter(Beverage == "Orange Juice") %>% 
  # Select
  select(Phylum, Class, Order, Family, Genus, Sample, Abundance, Treatment, Day, Trial, Shelf_Life_Stage, Spoilage) %>%
  filter(Abundance > 0) 
 
# Cider 16S data
filt_16S_cd <- genus_16S %>%
  filter(Beverage == "Cider") %>% 
  # Select
  select(Phylum, Class, Order, Family, Genus, Sample, Abundance, Treatment, Day, Trial, Shelf_Life_Stage, Spoilage) %>%
  filter(Abundance > 0) 

# OJ ITS data
filt_ITS_oj <- genus_ITS %>%
  filter(Beverage == "Orange Juice") %>% 
  # Select
  select(Phylum, Class, Order, Family, Genus, Sample, Abundance, Treatment, Day, Trial, Shelf_Life_Stage, Spoilage) %>%
  filter(Abundance > 0) 
 
# Cider ITS data
filt_ITS_cd <- genus_ITS %>%
  filter(Beverage == "Cider") %>% 
  # Select
  select(Phylum, Class, Order, Family, Genus, Sample, Abundance, Treatment, Day, Trial, Shelf_Life_Stage, Spoilage) %>%
  filter(Abundance > 0) 
```
### 3. Make compositional plots
```{r}
genus_16S_oj <- filt_16S_oj %>%
  select(Treatment, Trial, Shelf_Life_Stage, Spoilage, Genus, Abundance) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial) %>%
  mutate(totalSum = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus) %>%
  summarise(
    Abundance = sum(Abundance),
    totalSum,
    RelAb = Abundance/totalSum) %>%
  unique()

# Check summary statistics
max(genus_16S_oj$RelAb)
## [1] 0.9998634
mean(genus_16S_oj$RelAb)
## [1] 0.03408399
min(genus_16S_oj$RelAb)
## [1] 1.485619e-05
length(unique(genus_16S_oj$Genus))
## [1] 197
```
We cannot have a readable plot with 197 distinct colors, so we will have to further group some low-abundance taxa together.
```{r}
genus_16S_oj <- filt_16S_oj %>%
  select(Treatment, Trial, Shelf_Life_Stage, Spoilage, Genus, Abundance) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial) %>%
  mutate(totalSum = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%
  summarise(
    Abundance = sum(Abundance),
    Genus = ifelse(Abundance < 0.025, "Others", Genus)) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%  
# now group and summarize again to group newly labeled low abundance taxa together
  summarise(
    Abundance = sum(Abundance),
    RelAb = Abundance/totalSum) %>%
  unique()
# Now we are down to 20 genera (including "Others")
unique(genus_16S_oj$Genus)
```

# Set up color palette
```{r}
brewerPlus <- distinct_palette(pal = "brewerPlus")
scales::show_col(brewerPlus)

names(brewerPlus) <- unique(genus_16S_oj$Genus)
# Change 'Others' to grey
brewerPlus["Others"] <- "grey"
```

'Wrap' some treatment names that are too long
```{r}
Treatment_labs <- c("No Treatment",               "Pasteurization",          
"Pasteurization + \n SB + PS",
"Pasteurization + \n DMDC + NG",
"HPP", "HPP \n + DMDC + NG",           
"UV") 
names(Treatment_labs) <- c("No Treatment",               "Pasteurization",          
"Pasteurization + SB + PS",
"Pasteurization + DMDC + NG",
"HPP", "HPP + DMDC + NG",           
"UV") 
```
Now we are ready to plot
Bacterial communities in orange juice
```{r}
OJ_16S_compbar <- ggplot(genus_16S_oj)+
  geom_col(mapping = aes(x = Shelf_Life_Stage, y = RelAb, fill = Genus), color = "black", position = "stack", show.legend = TRUE)+
  facet_grid(cols = vars(Treatment), rows = vars(Trial), labeller = labeller(Treatment = Treatment_labs))+
  ylab("Relative Abundance") +
  xlab(NULL)+
  scale_fill_manual(values = brewerPlus) +
  theme_pubr()+
  theme(axis.text.y = element_text(size = 16, color = "black"),
        axis.title.y = element_text(size = 16, color = "black", face = "bold"),
        axis.text.x = element_text(size = 14, angle = 90, vjust = 1, hjust = 0.5, color = "black"),
        legend.text = element_text(size = 12, face = "italic"),
        legend.position = "bottom",
        legend.spacing.x = unit(0.1, 'mm'),
        legend.spacing.y = unit(0.05, 'mm'),
        plot.margin=grid::unit(c(0.1,0.1,0.1,0.1), "mm"),
        strip.text = element_text(face = "bold", angle = 0),
        legend.title = element_text(face="bold", size = 14))+
  guides(fill=guide_legend(ncol=3))
```

Fungal communities in orange juice
```{r}
genus_ITS_oj <- filt_ITS_oj %>%
  select(Treatment, Trial, Shelf_Life_Stage, Spoilage, Genus, Abundance) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial) %>%
  mutate(totalSum = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%
  summarise(
    Abundance = sum(Abundance),
    Genus = ifelse(Abundance < 0.025, "Others", Genus)) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%  
# now group and summarize again to group newly labeled low abundance taxa together
  summarise(
    Abundance = sum(Abundance),
    RelAb = Abundance/totalSum) %>%
  unique()
# Now we are down to 22 genera (including "Others")
unique(genus_ITS_oj$Genus)
```

```{r}
kelly <- microViz::distinct_palette(pal = "kelly", add = "chartreuse")
kelly <- append(kelly,"deeppink1")
names(kelly) <- unique(genus_ITS_oj$Genus)
# Assign grey to 'Others'
kelly["Others"] <- "lightgrey"

OJ_ITS_compbar <- ggplot(genus_ITS_oj)+
  geom_col(mapping = aes(x = Shelf_Life_Stage, y = RelAb, fill = Genus), color = "black", position = "stack", show.legend = TRUE)+
  facet_grid(cols = vars(Treatment), rows = vars(Trial), labeller = labeller(Treatment = Treatment_labs))+
  ylab("Relative Abundance") +
  xlab(NULL)+
  scale_fill_manual(values = kelly) +
  theme_pubr()+
  theme(axis.text.y = element_text(size = 16, color = "black"),
        axis.title.y = element_text(size = 16, color = "black", face = "bold"),
        axis.text.x = element_text(size = 14, angle = 90, vjust = 1, hjust = 0.5, color = "black"),
        legend.text = element_text(size = 12, face = "italic"),
        legend.position = "bottom",
        legend.spacing.x = unit(0.1, 'mm'),
        legend.spacing.y = unit(0.05, 'mm'),
        plot.margin=grid::unit(c(0.1,0.1,0.1,0.1), "mm"),
        strip.text = element_text(face = "bold", angle = 0),
        legend.title = element_text(face="bold", size = 14))+
  guides(fill=guide_legend(ncol=3))
```

Display the combined 16S and ITS plot for OJ
```{r}
ggarrange(OJ_16S_compbar, OJ_ITS_compbar, nrow = 1, labels = "AUTO")

```
Save plot
```{r eval=FALSE, include=FALSE}
ggsave("Figure/OJ_compositional.png", device = "png", height = 9, width = 13, units = "in")

```
Bacterial communities in cider
```{r}
genus_16S_cd <- filt_16S_cd %>%
  select(Treatment, Trial, Shelf_Life_Stage, Spoilage, Genus, Abundance) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial) %>%
  mutate(totalSum = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%
  summarise(
    Abundance = sum(Abundance),
    Genus = ifelse(Abundance < 0.05, "Others", Genus)) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%  
# now group and summarize again to group newly labeled low abundance taxa together
  summarise(
    Abundance = sum(Abundance),
    RelAb = Abundance/totalSum) %>%
  unique()
# Now we are down to 20 genera (including "Others")
unique(genus_16S_cd$Genus)
genus_16S_cd$Genus = gsub("Methylobacterium-Methylorubrum", "Methylobacterium- \n Methylorubrum", genus_16S_cd$Genus)
```

```{r}
brewerPlus <- microViz::distinct_palette(pal = "brewerPlus", add = NA)
names(brewerPlus) <- unique(genus_16S_cd$Genus)
# Change 'Others' to grey
brewerPlus["Others"] <- "grey"

CD_16S_compbar <- ggplot(genus_16S_cd)+
  geom_col(mapping = aes(x = Shelf_Life_Stage, y = RelAb, fill = Genus), color = "black", position = "stack", show.legend = TRUE)+
  facet_grid(cols = vars(Treatment), rows = vars(Trial), labeller = labeller(Treatment = Treatment_labs))+
  ylab("Relative Abundance") +
  xlab(NULL)+
  scale_fill_manual(values = brewerPlus) +
  theme_pubr()+
  theme(axis.text.y = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 18, color = "black", face = "bold"),
        axis.text.x = element_text(size = 18, angle = 90, vjust = 1, hjust = 0.5, color = "black"),
        legend.text = element_text(size = 13, face = "italic"),
        legend.position = "bottom",
        legend.spacing.x = unit(0.1, 'mm'),
        legend.spacing.y = unit(0.05, 'mm'),
        plot.margin=grid::unit(c(0.1,0.1,0.1,0.1), "mm"),
        strip.text = element_text(face = "bold", angle = 0),
        legend.title = element_text(face="bold", size = 18))+
  guides(fill=guide_legend(ncol=3))
```

Fungal communities in cider
```{r}
genus_ITS_cd <- filt_ITS_cd %>%
  select(Treatment, Trial, Shelf_Life_Stage, Spoilage, Genus, Abundance) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial) %>%
  mutate(totalSum = sum(Abundance)) %>%
  ungroup() %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%
  summarise(
    Abundance = sum(Abundance),
    Genus = ifelse(Abundance < 0.05, "Others", Genus)) %>%
  group_by(Treatment, Shelf_Life_Stage, Trial, Genus, totalSum) %>%  
# now group and summarize again to group newly labeled low abundance taxa together
  summarise(
    Abundance = sum(Abundance),
    RelAb = Abundance/totalSum) %>%
  unique()
# Now we are down to 19 genera (including "Others")
unique(genus_ITS_cd$Genus)
```

```{r}
kelly <- microViz::distinct_palette(pal = "kelly")
names(kelly) <- unique(genus_ITS_cd$Genus)
kelly["Others"] <- "lightgrey"

CD_ITS_compbar <-
ggplot(genus_ITS_cd)+
  geom_col(mapping = aes(x = Shelf_Life_Stage, y = RelAb, fill = Genus), color = "black", position = "stack", show.legend = TRUE)+
  facet_grid(cols = vars(Treatment), rows = vars(Trial), labeller = labeller(Treatment = Treatment_labs))+
  ylab("Relative Abundance") +
  xlab(NULL)+
  scale_fill_manual(values = kelly) +
  theme_pubr()+
  theme(axis.text.y = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 18, color = "black", face = "bold"),
        axis.text.x = element_text(size = 18, angle = 90, vjust = 1, hjust = 0.5, color = "black"),
        legend.text = element_text(size = 13, face = "italic"),
        legend.position = "bottom",
        legend.spacing.x = unit(0.1, 'mm'),
        legend.spacing.y = unit(0.05, 'mm'),
        plot.margin=grid::unit(c(0.1,0.1,0.1,0.1), "mm"),
        strip.text = element_text(face = "bold", angle = 0),
        legend.title = element_text(face="bold", size = 18))+
  guides(fill=guide_legend(ncol=3))
```

```{r}
ggarrange(CD_16S_compbar, CD_ITS_compbar, ncol = 2, labels = "AUTO")

```
Save plot
```{r eval=FALSE, include=FALSE}
ggsave("Figure/CD_compositional.png", device = "png", height = 9, width = 13, units = "in")

```